---
title: "allofus-HIV-cohort-definitions"
output: html_document
date: "2026-01-08"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE)
```


# Load packages
```{r}
# install.packages(c("tidyverse", "allofus", "gt", "ggvenn", "ggh4x", "ggplot2"))

library(allofus)
library(tidyverse)
library(gt)
library(ggvenn)
library(ggh4x)
library(ggplot2)
```

# Connecting to the allofus database
```{r connection}
con <- allofus::aou_connect()
```
# Once connected you can begin to Create the  Cohorts
# Creating HIV cohorts using survey data, condition_occurrence data, and drug_data



#Survey Cohort

# Have you ever been diagnosed with the following conditions? - HIV/AIDS
# https://athena.ohdsi.org/search-terms/terms/1384391
```{r}
svy = aou_survey(questions = 1384391) %>%
    filter(infectiousdiseasecondition_hivaids_yes == "Yes") %>% # only Yes responses
    select(person_id) %>%
    mutate(svy = "s") # create a variable indicating that it was from the survey data.

# preview results
tally(svy) # size of cohort
```


## Condition cohort ##

# Code to extract people with HIV (except HIV negative (code is above))
# Concept IDs in the vector(where are they found)
# Human immunodeficiency virus carrier(pheno lib + AoU Data brow), Human immunodeficiency virus II infection(pheno lib + AoU Data brow), HIV positive(AoU Data brow), Human immunodeficiency virus infection(pheno lib + AoU Data brow)
# https://data.ohdsi.org/PhenotypeLibrary/
```{r}
hiv_ancestors <- c(44783356, 4013106, 432554, 439727)
hiv_cond_concepts <- tbl(con, "concept_ancestor") %>%
    filter(ancestor_concept_id %in% hiv_ancestors) %>%
    distinct(descendant_concept_id)

# confirmed ancestors are included
hiv_cond_concepts %>% filter(descendant_concept_id %in% hiv_ancestors)

#Code to extract people with HIV Concept ID + descendant
cond <- tbl(con, "condition_occurrence") %>%
    inner_join(hiv_cond_concepts, by = join_by(condition_concept_id == descendant_concept_id)) %>%
    distinct(person_id) %>%
    mutate(cond = "c") # from the condition table
tally(cond) # size of cohort
```


## Drug cohort ##

# Drug list was developed from the HIV drugs and then by mapping to all descendents and
# examined by hand with Chelsea/Brianne to make sure it includes only desired drugs
# Saved file to github to be read into allofus
# Load drug lists
```{r}
HIV_Drugs <- read.csv("https://raw.githubusercontent.com/plummern/EHR_DRUGS/main/HIV_Drugs_Descendants_FINAL.csv")
new <- read.csv("https://raw.githubusercontent.com/plummern/AoU_EHR_data/refs/heads/main/NEW%20method%20with%20J05_AoU%20drugs%20compared%20with%20Brianne's%20HIV%20drug%20list.csv")
reviewed_drug <- read.csv("https://raw.githubusercontent.com/plummern/AoU_EHR_data/refs/heads/main/drugs_only_in_new_BOM_REVIEWED.csv")

# Standardize concept_code type across all
HIV_Drugs <- HIV_Drugs %>% mutate(concept_code = as.character(concept_code))
new <- new %>% mutate(concept_code = as.character(concept_code))
reviewed_drug <- reviewed_drug %>% mutate(concept_code = as.character(concept_code))

# Combine lists
combined_drugs <- bind_rows(HIV_Drugs, new, reviewed_drug)

# Filter final drug list
drugs <- combined_drugs %>%
  filter(drugs_to_keep == 1) %>%
  select(concept_id) %>%
  distinct()

# Temp table for join
hiv_drug_concepts <- allofus::aou_create_temp_table(data = drugs)

# Get unique people with matching HIV drugs
drug <- tbl(con, "drug_exposure") %>%
  inner_join(hiv_drug_concepts, by = join_by(drug_concept_id == concept_id)) %>%
  distinct(person_id) %>%
  mutate(drug = "d")

# Count how many individuals matched
drug %>% tally()
```


## Viral Load (VL) cohort ##
```{r}
loincs <- tbl(con, "concept") %>% 
    filter(domain_id == "Measurement",
           vocabulary_id == "LOINC",
           concept_class_id == "Lab Test") %>% 
    collect()
 
hivs_VL <- loincs %>% 
    filter(str_detect(concept_name, "^HIV"),
           str_detect(concept_name, "viral load"))

# Extract relevant concept_ids for HIV viral load measurements and Remove duplicate person_ids from the vl dataset
VL_concept_ids <- hivs_VL$concept_id

vl <- tbl(con, "measurement") %>%
    filter(measurement_concept_id %in% VL_concept_ids) %>% 
    inner_join(tbl(con, "concept"), by = c("measurement_concept_id" = "concept_id")) %>%
    distinct(person_id, .keep_all = TRUE) %>%  # Keep only one entry per person_id
    distinct(person_id) %>%
    mutate(vl = "v")

# View the cleaned dataset without NA values in value_as_number
tally(vl)
```


## Join Together ##
```{r}
# NA values will indicate that the person_id did not show up for that cohort. 
all = full_join(svy, cond, by = join_by("person_id")) %>%
    full_join(drug, by = join_by("person_id")) %>%
    full_join(vl, by = join_by("person_id")) %>%
    collect()
# preview increased from 9,299 to 9,463 once viral load was joined to the HIV cohort
tally(all)
```

# replace NA with blanks so that we can combined the columns into the combination of s, c, d, vl
# then change NA values to 0
```{r}
cohorts = all %>% 
    replace_na(list(svy = "", cond = "", drug = "", vl = "")) %>%  # Replace NA with blanks
    unite(cohorts, c("svy", "cond", "drug", "vl"), sep = "", remove = FALSE) %>%  # Combine columns into one          
    mutate(across(c("svy", "cond", "drug", "vl"), ~ifelse(!is.na(.x) & !.x=="", 1, 0)))
count(cohorts, cohorts)
```
# Adding Demographics
```{r}
# Write the cohorts table to a temp table to use on the db level
all_tbl = allofus::aou_create_temp_table(cohorts, nchar_batch = 1e+04)
# Demographic variables from chelsea. We'll only use a subset of these

demo <- read.csv("https://raw.githubusercontent.com/cnwong/AoU_FI/main/demo_vars.csv")
#Applying the aou_surevey demo info to 
aou_data <- aou_survey(
    all_tbl,
    questions = demo$concept_id,
    question_output = demo$demo,)
```

# age_at_consent from cb_search_person is actually the date individuals consented to contribute EHR data
# its not the age at which they enrolled in the study
# see https://docs.google.com/spreadsheets/d/1XLVq84LLd0VZMioF2sPwyiaPw3EFp5c8o1CTWGPH-Yc/edit?pli=1#gid=183931508
```{r}
aou_data_demo = aou_data %>%
  mutate(
      race_cat = case_when(
          race == "White" ~ "White",
          race == "Black" ~ "Black",
          race == "Hispanic" ~ "Hispanic/Latino",
          race %in% c("Asian","NHPI","MENA","RaceEthnicityNoneOfThese") ~ "Other",
          #race == "MENA" ~ "Other",
          #race ==  ~ "Other",
          race %in% c("Skip", "PreferNotToAnswer") ~ "Skip",
          race %in% c("Hispanic,MENA,White","Black,NHPI","Hispanic,White","Black,White","Asian,White",
                     "MENA,White","Asian,Hispanic", "Black,Hispanic","Hispanic,MENA", "NHPI,White", "Asian,Black",
                     "Black,Hispanic,MENA,White","Hispanic,NHPI,White","Black,Hispanic,MENA,NHPI","Asian,MENA,White",
                     "Black,Hispanic,White","Asian,Black,Hispanic,White","Asian,Hispanic,White","Black,MENA", "Asian,Black,White",
                     "Asian,NHPI,White","Black,Hispanic,MENA","Asian,MENA","Asian,Black,NHPI", "Black,MENA,White", "Hispanic,NHPI",
                     "MENA,NHPI,White","Black,NHPI,White","Asian,NHPI","Asian,Black,MENA,White","Asian,Hispanic,MENA,White",
                     "Black,MENA,NHPI","Asian,Black,NHPI,White", "Asian,Black,Hispanic", "Asian,Black,Hispanic,MENA,White", 
                     "Asian,Hispanic,NHPI,White","Hispanic,MENA,NHPI", "Asian,Hispanic,MENA,NHPI", "MENA,NHPI", "Black,MENA,NHPI,White",
                     "Asian,Hispanic,NHPI","Asian,Black,Hispanic,NHPI,White","Black,Hispanic,NHPI","Asian,Black,Hispanic,MENA",
                     "Black,Hispanic,NHPI,White","Asian,Black,MENA") ~ "Mixed"
     ),
  sex_birth_cat = case_when(
              sex_birth == "Male" ~ "Male",
              sex_birth == "Female" ~ "Female",
              sex_birth %in% c ("Intersex","SexAtBirthNoneOfThese") ~ "Diverse",
              sex_birth %in% c("Skip", "PreferNotToAnswer") ~ "Skip",
            TRUE ~ NA_character_
        ),
        sex_birth_bin = case_when(
              sex_birth == "Male" ~ 0,
              sex_birth == "Female" ~ 1,
              sex_birth %in% c ("Intersex","SexAtBirthNoneOfThese") ~ 2,
              sex_birth %in% c("Skip", "PreferNotToAnswer") ~  NA_real_,
              TRUE ~ NA_real_
        ),
        gender_cat = case_when(
              gender=="Man" ~ "Man",
              gender == "Woman" ~ "Woman",
              gender %in% c("NonBinary","AdditionalOptions","Transgender","NonBinary,Woman", "Man,NonBinary",
                        "Man,Transgender","Man,Woman","Transgender,Woman","AdditionalOptions,Woman",
                        "Man,NonBinary,Transgender,Woman","NonBinary,Transgender","NonBinary,Transgender,Woman",
                        "Man,NonBinary,Woman","AdditionalOptions,Man","Man,Transgender,Woman") ~ "Diverse",
              gender %in% c("Skip", "PreferNotToAnswer") ~ "Skip",
            TRUE ~ NA_character_
        ),
          gender_bin = case_when(
              gender=="Man" ~ 0,
              gender == "Woman" ~ 1,
              gender %in% c("NonBinary","AdditionalOptions","Transgender","NonBinary,Woman", "Man,NonBinary",
                                "Man,Transgender","Man,Woman","Transgender,Woman","AdditionalOptions,Woman",
                                "Man,NonBinary,Transgender,Woman","NonBinary,Transgender","NonBinary,Transgender,Woman",
                                "Man,NonBinary,Woman","AdditionalOptions,Man","Man,Transgender,Woman") ~ 2,
              gender %in% c("Skip", "PreferNotToAnswer") ~  NA_real_,
              TRUE ~ NA_real_
        ),
  gender_sex_cat = case_when(
      is.na(sex_birth_cat) & is.na(gender_cat) ~ NA_character_,
      gender_cat == "Diverse" ~ "Gender Diverse",
      sex_birth_cat == "Diverse" ~ "Gender Diverse",
      sex_birth_cat == "Male" & gender_cat == "Woman" ~ "Gender Diverse",
      sex_birth_cat == "Female" & gender_cat == "Man" ~ "Gender Diverse",
      sex_birth_cat == "Male" & gender_cat == "Man" ~ "Male",
      sex_birth_cat == "Female" & gender_cat == "Woman" ~ "Female",
      sex_birth_cat == "Skip" | gender_cat == "Skip" ~ NA_character_)) %>%
distinct( # we only need these columns
    person_id,
    race, race_cat,
    sex_birth, sex_birth_cat, sex_birth_bin,
    gender,gender_cat, gender_bin,
    gender_sex_cat
) %>%
    # a not ideal way to get age. 
  left_join(tbl(con, "cb_search_person") %>% select(person_id, age_at_consent, has_ehr_data),
            by = join_by(person_id))
```

# edited the age groups after reviewing the national data
# this would change the levels to be what we want to see
# appending the name of the data with _c to indicate that it's been collected. 

# Collecting demographic data and processing it
```{r}
aou_data_demo_c <- collect(aou_data_demo) %>%
    rename(race_eth_cat = race_cat) %>%  # Rename race_cat to race_eth_cat
    mutate(
        age_cat = cut(as.numeric(age_at_consent),  # Categorize age into ranges
                      breaks = c(12, 24, 34, 44, 54, 92),
                      labels = c("Age 13-24", "Age 25-34", "Age 35-44", "Age 45-54", "Age 55+")),
        age_cat = fct_relevel(age_cat, "Age 13-24", "Age 25-34", "Age 35-44", "Age 45-54", "Age 55+"),
        age_flag_65_plus = ifelse(age_at_consent >= 65, "65+", "Under 65"),    # Flag for age 65 and older
        age_flag_65_plus = fct_relevel(age_flag_65_plus, "Under 65", "65+"),
        gender_sex_cat = fct_relevel(gender_sex_cat, "Male", "Female", "Gender Diverse"),  # Reorder gender categories
        race_eth_cat = fct_relevel(race_eth_cat,  # Reorder race/ethnicity categories
                                   "White", "Black", "Hispanic/Latino", "Mixed", "Other", "Skip")
    ) %>%
    select(person_id, age_cat, gender_sex_cat, race_eth_cat, age_at_consent, age_flag_65_plus)  # Select relevant columns
```

# join the demographic data to the cohorts table 
# change all character columns to factor
# preview data to see total sample size.(n = 9,299 7/16/24).. new:(n = 9,998 12/12/24).. newest:(n = 10,022 6/23/25)
```{r}
demo_clean = cohorts %>%
    left_join(aou_data_demo_c, by = "person_id") %>%
    mutate(across(where(is.character), as.factor)) %>%
    select(person_id, cohorts, svy, cond, drug, vl, 
          gender_sex_cat, race_eth_cat, age_cat, age_flag_65_plus)
```

# Making Summary Data

# getting the counts of each combination for reference later on
```{r}
demo_clean %>%
    summarise(
    # Individual sources
    svy = sum(str_detect(cohorts, "s")),
    cond = sum(str_detect(cohorts, "c")),
    drug = sum(str_detect(cohorts, "d")),
    vl = sum(str_detect(cohorts, "v")),

    # Specific combinations
    svy_and_cond_and_drug_and_vl = sum(
      str_detect(cohorts, "s") & str_detect(cohorts, "c") & str_detect(cohorts, "d") & str_detect(cohorts, "v")
    ),

    svy_and_cond_and_drug_or_vl = sum(
      str_detect(cohorts, "s") & str_detect(cohorts, "c") & (str_detect(cohorts, "d") | str_detect(cohorts, "v"))
    ),

    svy_and_vl_and_cond_or_drug = sum(
      str_detect(cohorts, "s") & str_detect(cohorts, "v") & (str_detect(cohorts, "c") | str_detect(cohorts, "d"))
    ),

    drug_and_vl_and_svy_or_cond = sum(
      str_detect(cohorts, "d") & str_detect(cohorts, "v") & (str_detect(cohorts, "s") | str_detect(cohorts, "c"))
    ),

    svy_and_cond_or_drug = sum(
      str_detect(cohorts, "s") & (str_detect(cohorts, "c") | str_detect(cohorts, "d"))
    ),

    vl_and_cond_or_drug = sum(
      str_detect(cohorts, "v") & (str_detect(cohorts, "c") | str_detect(cohorts, "d"))
    ),

    vl_and_svy_or_drug = sum(
      str_detect(cohorts, "v") & (str_detect(cohorts, "s") | str_detect(cohorts, "d"))
    ),

    vl_and_svy_or_cond = sum(
      str_detect(cohorts, "v") & (str_detect(cohorts, "s") | str_detect(cohorts, "c"))
    ),

    svy_or_cond_or_drug_or_vl = sum(
      str_detect(cohorts, "s|c|d|v")
    ),

    # 1-to-1 pairs and other combinations
    cond_and_drug = sum(
      str_detect(cohorts, "c") & str_detect(cohorts, "d")
    ),

    svy_and_cond = sum(
      str_detect(cohorts, "s") & str_detect(cohorts, "c")
    ),

    svy_and_drug = sum(
      str_detect(cohorts, "s") & str_detect(cohorts, "d")
    ),

    vl_and_svy = sum(
      str_detect(cohorts, "v") & str_detect(cohorts, "s")
    ),

    vl_and_cond = sum(
      str_detect(cohorts, "v") & str_detect(cohorts, "c")
    ),

    vl_and_drug = sum(
      str_detect(cohorts, "v") & str_detect(cohorts, "d")))
```


# We need the size of each cell to calculate proprtion representation within each later on
# needs to be in long format too

# unique combinations - these are the 'and' denominators 
```{r}
ands = demo_clean %>% count(cohorts, name = "total" )

# alls and ors
alls_ors = demo_clean %>%
    summarize(svy = sum(str_detect(cohorts, "s")),
             cond = sum(str_detect(cohorts, "c")),
             drug = sum(str_detect(cohorts, "d")),
             vl = sum(str_detect(cohorts, "v")),
             svy_and_cond_or_drug = sum(
                str_detect(cohorts, "s") & str_detect(cohorts, "c|d")
            ),
            svy_and_cond = sum(
                str_detect(cohorts, "s") & str_detect(cohorts, "c")
            ),
            svy_and_drug = sum(
                str_detect(cohorts, "s") & str_detect(cohorts, "d")
            ),
            cond_and_drug = sum(
                str_detect(cohorts, "c") & str_detect(cohorts, "d")
            ),
            svy_and_cond_and_drug_or_vl = sum(
                str_detect(cohorts, "s") & str_detect(cohorts, "c") & str_detect(cohorts, "d|v")
            ),
            svy_and_vl_and_cond_or_drug = sum(
                str_detect(cohorts, "s") & str_detect(cohorts, "v") & str_detect(cohorts, "c|d")
            ),
            drug_and_vl_and_svy_or_cond = sum(
                 str_detect(cohorts, "d") & str_detect(cohorts, "v") & str_detect(cohorts, "s|c")
            ),
            vl_and_cond_or_drug = sum(
                str_detect(cohorts, "v") & str_detect(cohorts, "c|d")
            ),
            vl_and_svy_or_drug = sum(
                str_detect(cohorts, "v") & str_detect(cohorts, "s|d")
            ),
            vl_and_svy_or_cond = sum(
                str_detect(cohorts, "v") & str_detect(cohorts, "s|c")
            ),
            vl_and_svy = sum(
                str_detect(cohorts, "v") & str_detect(cohorts, "s")
            ),
            vl_and_cond = sum(
                str_detect(cohorts, "v") & str_detect(cohorts, "c")
            ),
            vl_and_drug = sum(
                str_detect(cohorts, "v") & str_detect(cohorts, "d")
            ),
            svy_or_cond_or_drug_or_vl = sum(
                    str_detect(cohorts, "s|c|d|v")
                )) %>%
    pivot_longer(cols = 1:18, names_to = "cohorts", values_to = "total")

nat = tibble(cohorts = "nat", total = 1086806) # nat = national data statistics


denominators = 
    bind_rows(
      alls_ors, ands, nat)
```


# now we need to calcualte demographics for each cohort/source combination we're interested in
# first the "and" conditions
# then the next cell has the "or" conditions

# A bunch of clean up to calculate cell sizes, rename items, and join to denominators
# done for the ands and the alls/ors separately and then combined. 

# ands
```{r}
age = demo_clean %>%
    count(cohorts, age_cat) %>%
    left_join(denominators, by = join_by(cohorts)) %>%
    mutate(variable = "age")%>%
    rename("cat" = ends_with("cat"))

gender_sex = demo_clean %>%
    count(cohorts, gender_sex_cat) %>%
    left_join(denominators, by = join_by(cohorts)) %>%
    mutate(variable = "gender_sex")%>%
    rename("cat" = ends_with("cat"))


race = demo_clean %>%
    count(cohorts, race_eth_cat) %>%
    left_join(denominators, by = join_by(cohorts)) %>%
    mutate(variable = "race")%>%
    rename("cat" = ends_with("cat"))

# example preview output
head(age)
```


# anys ors
```{r}
age_any = demo_clean %>%
    mutate(
            svy_and_cond_or_drug = ifelse(
                str_detect(cohorts, "s") & str_detect(cohorts, "c|d"), 1, 0
            ),
            svy_and_cond = ifelse(
                str_detect(cohorts, "s") & str_detect(cohorts, "c"), 1, 0
            ),
            svy_and_drug = ifelse(
                str_detect(cohorts, "s") & str_detect(cohorts, "d"), 1, 0
            ),
            cond_and_drug = ifelse(
                str_detect(cohorts, "c") & str_detect(cohorts, "d"), 1, 0
            ),
            svy_and_cond_and_drug_or_vl = ifelse(
                str_detect(cohorts, "s") & str_detect(cohorts, "c") & str_detect(cohorts, "d|v"), 1, 0
            ),
            svy_and_vl_and_cond_or_drug = ifelse(
                str_detect(cohorts, "s") & str_detect(cohorts, "v") & str_detect(cohorts, "c|d"), 1, 0
            ),
            drug_and_vl_and_svy_or_cond = ifelse(
                 str_detect(cohorts, "d") & str_detect(cohorts, "v") & str_detect(cohorts, "s|c"), 1, 0
            ),
            vl_and_cond_or_drug = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "c|d"), 1, 0
            ),
            vl_and_svy_or_drug = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "s|d"), 1, 0
            ),
            vl_and_svy_or_cond = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "s|c"), 1, 0
            ),
            vl_and_svy = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "s"), 1, 0
            ),
            vl_and_cond = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "c"), 1, 0
            ),
            vl_and_drug = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "d"), 1, 0
            ),
            svy_or_cond_or_drug_or_vl = ifelse(
                    str_detect(cohorts, "s|c|d|v"), 1, 0
                )) %>%
    select(person_id, svy:vl,
           svy_and_cond_or_drug,
           svy_and_cond, svy_and_drug, cond_and_drug,
           svy_and_cond_and_drug_or_vl, svy_and_vl_and_cond_or_drug, drug_and_vl_and_svy_or_cond,
           vl_and_cond_or_drug, vl_and_svy_or_drug, vl_and_svy_or_cond,
           vl_and_svy, vl_and_cond, vl_and_drug,
           svy_or_cond_or_drug_or_vl,
           ends_with("cat")) %>%
    pivot_longer(cols = svy:svy_or_cond_or_drug_or_vl, names_to = "cohorts") %>%
    filter(value == 1) %>%
    count(age_cat, cohorts) %>%
    left_join(denominators, by = join_by(cohorts)) %>%
    mutate(cohorts = case_when(
        cohorts=="cond" ~ "any_c",
        cohorts=="svy" ~ "any_s",
        cohorts=="drug" ~ "any_d",
        cohorts=="vl" ~ "any_v",
        TRUE ~ cohorts
    
    ),
        variable = "age"
    ) %>%
    rename(cat = age_cat)

gender_sex_any = demo_clean %>%
    mutate(
            svy_and_cond_or_drug = ifelse(
                str_detect(cohorts, "s") & str_detect(cohorts, "c|d"), 1, 0
            ),
            svy_and_cond = ifelse(
                str_detect(cohorts, "s") & str_detect(cohorts, "c"), 1, 0
            ),
            svy_and_drug = ifelse(
                str_detect(cohorts, "s") & str_detect(cohorts, "d"), 1, 0
            ),
            cond_and_drug = ifelse(
                str_detect(cohorts, "c") & str_detect(cohorts, "d"), 1, 0
            ),
            svy_and_cond_and_drug_or_vl = ifelse(
                str_detect(cohorts, "s") & str_detect(cohorts, "c") & str_detect(cohorts, "d|v"), 1, 0
            ),
            svy_and_vl_and_cond_or_drug = ifelse(
                str_detect(cohorts, "s") & str_detect(cohorts, "v") & str_detect(cohorts, "c|d"), 1, 0
            ),
            drug_and_vl_and_svy_or_cond = ifelse(
                 str_detect(cohorts, "d") & str_detect(cohorts, "v") & str_detect(cohorts, "s|c"), 1, 0
            ),
            vl_and_cond_or_drug = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "c|d"), 1, 0
            ),
            vl_and_svy_or_drug = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "s|d"), 1, 0
            ),
            vl_and_svy_or_cond = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "s|c"), 1, 0
            ),
            vl_and_svy = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "s"), 1, 0
            ),
            vl_and_cond = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "c"), 1, 0
            ),
            vl_and_drug = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "d"), 1, 0
            ),
            svy_or_cond_or_drug_or_vl = ifelse(
                    str_detect(cohorts, "s|c|d|v"), 1, 0
                )) %>%
    select(person_id, svy:vl,
           svy_and_cond_or_drug,
           svy_and_cond, svy_and_drug, cond_and_drug,
           svy_and_cond_and_drug_or_vl, svy_and_vl_and_cond_or_drug, drug_and_vl_and_svy_or_cond,
           vl_and_cond_or_drug, vl_and_svy_or_drug, vl_and_svy_or_cond,
           vl_and_svy, vl_and_cond, vl_and_drug,
           svy_or_cond_or_drug_or_vl,
           ends_with("cat")) %>%
    pivot_longer(cols = svy:svy_or_cond_or_drug_or_vl, names_to = "cohorts") %>%
    filter(value == 1) %>%
    count(gender_sex_cat, cohorts) %>%
    left_join(denominators, by = join_by(cohorts)) %>%
    mutate(cohorts = case_when(
        cohorts=="cond" ~ "any_c",
        cohorts=="svy" ~ "any_s",
        cohorts=="drug" ~ "any_d",
        cohorts=="vl" ~ "any_v",
        TRUE ~ cohorts
    
    ),
        variable = "gender_sex"
    ) %>%
    rename(cat = gender_sex_cat)


race_any = demo_clean %>%
    mutate(
            svy_and_cond_or_drug = ifelse(
                str_detect(cohorts, "s") & str_detect(cohorts, "c|d"), 1, 0
            ),
            svy_and_cond = ifelse(
                str_detect(cohorts, "s") & str_detect(cohorts, "c"), 1, 0
            ),
            svy_and_drug = ifelse(
                str_detect(cohorts, "s") & str_detect(cohorts, "d"), 1, 0
            ),
            cond_and_drug = ifelse(
                str_detect(cohorts, "c") & str_detect(cohorts, "d"), 1, 0
            ),
            svy_and_cond_and_drug_or_vl = ifelse(
                str_detect(cohorts, "s") & str_detect(cohorts, "c") & str_detect(cohorts, "d|v"), 1, 0
            ),
            svy_and_vl_and_cond_or_drug = ifelse(
                str_detect(cohorts, "s") & str_detect(cohorts, "v") & str_detect(cohorts, "c|d"), 1, 0
            ),
            drug_and_vl_and_svy_or_cond = ifelse(
                 str_detect(cohorts, "d") & str_detect(cohorts, "v") & str_detect(cohorts, "s|c"), 1, 0
            ),
            vl_and_cond_or_drug = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "c|d"), 1, 0
            ),
            vl_and_svy_or_drug = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "s|d"), 1, 0
            ),
            vl_and_svy_or_cond = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "s|c"), 1, 0
            ),
            vl_and_svy = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "s"), 1, 0
            ),
            vl_and_cond = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "c"), 1, 0
            ),
            vl_and_drug = ifelse(
                str_detect(cohorts, "v") & str_detect(cohorts, "d"), 1, 0
            ),
            svy_or_cond_or_drug_or_vl = ifelse(
                    str_detect(cohorts, "s|c|d|v"), 1, 0
                )) %>%
    select(person_id, svy:vl,
           svy_and_cond_or_drug,
           svy_and_cond, svy_and_drug, cond_and_drug,
           svy_and_cond_and_drug_or_vl, svy_and_vl_and_cond_or_drug, drug_and_vl_and_svy_or_cond,
           vl_and_cond_or_drug, vl_and_svy_or_drug, vl_and_svy_or_cond,
           vl_and_svy, vl_and_cond, vl_and_drug,
           svy_or_cond_or_drug_or_vl,
           ends_with("cat")) %>%
    pivot_longer(cols = svy:svy_or_cond_or_drug_or_vl, names_to = "cohorts") %>%
    filter(value == 1) %>%
    count(race_eth_cat, cohorts) %>%
    left_join(denominators, by = join_by(cohorts)) %>%
    mutate(cohorts = case_when(
        cohorts=="cond" ~ "any_c",
        cohorts=="svy" ~ "any_s",
        cohorts=="drug" ~ "any_d",
        cohorts=="vl" ~ "any_v",
        TRUE ~ cohorts
    
    ),
        variable = "race"
    ) %>%
    rename(cat = race_eth_cat)
```

# combine them all together (stack them)
```{r}
aou_dat = bind_rows(age, age_any, gender_sex, gender_sex_any, race, race_any) %>%
    mutate(prop = n/total)

head(aou_dat)
```

# creating a dataframe for National data, matching the variables and categories of the AOU HIV data

# Other variable is the sum of prevalence (cases) for Asian + American Indian/Alaska Native + Native Hawaiian/Other Pacific Islander
# Data comes from here: https://aidsvu.org/resources/#/datasets?years=2021
# AidsVu uses CDC data with some changes I'm not quite sure about
```{r}
nat_dat <- tribble (
    ~variable, ~cat, ~n, ~total,
    "gender_sex", "Male", 836283, 1086806,
    "gender_sex", "Female", 250523, 1086806,
    "gender_sex", "Skip", 0, 1086806,
    "race", "Black", 432249, 1086806,
    "race", "White", 306474, 1086806,
    "race", "Hispanic/Latino", 270823, 1086806,
    "race", "Mixed", 55723, 1086806,
    "race", "Other", 20858, 1086806,
    "race", "Skip", 0, 1086806,
    "age", "Age 13-24", 28255, 1086806,
    "age", "Age 25-34", 164890, 1086806,
    "age", "Age 35-44", 205530, 1086806,
    "age", "Age 45-54", 246872, 1086806,
    "age", "Age 55+", 441259, 1086806,
) |>
    mutate(prop = n/total, cohorts = "nat")

# get percentages to verify later
nat_dat %>% group_by(variable, cat) %>% summarize(tot = sum(prop))

# put the aou and national data together
plot_dat = bind_rows(aou_dat, nat_dat)

# check unique condition counts
plot_dat %>% count(cohorts)
# cell sizes here are ok. A few NA stragglers that are ignorable
# sd has very small cell counts and is missing some
```


# Creating Proportion demographics by cohort

# Cleaning up variable names for the plot
```{r}
study_dat = plot_dat %>%
    filter(cohorts %in% c("nat", "scdv", "svy_and_cond_and_drug_or_vl", "svy_and_vl_and_cond_or_drug", "drug_and_vl_and_svy_or_cond",
                          "svy_and_cond_or_drug", "vl_and_cond_or_drug", "vl_and_svy_or_drug", "vl_and_svy_or_cond",
                          "svy_or_cond_or_drug_or_vl", "any_s", "any_c", "any_d", "any_v",
                          "cond_and_drug", "svy_and_cond", "svy_and_drug", "vl_and_svy", "vl_and_cond", "vl_and_drug")) %>%
    mutate(cohort_name = case_when(
        cohorts == "nat" ~ "national",
        cohorts == "scdv" ~ "surv & con & drug & vl",
        cohorts == "svy_and_cond_and_drug_or_vl" ~ "surv & con & drug | vl",
        cohorts == "svy_and_vl_and_cond_or_drug" ~ "surv & vl & con | drug",
        cohorts == "drug_and_vl_and_svy_or_cond" ~ "drug & vl & surv | con",
        cohorts == "svy_and_cond_or_drug" ~ "surv & con | drug",
        cohorts == "vl_and_cond_or_drug" ~ "vl & con | drug",
        cohorts == "vl_and_svy_or_drug" ~ "vl & surv | drug",
        cohorts == "vl_and_svy_or_cond" ~ "vl & surv | con",
        cohorts == "svy_or_cond_or_drug_or_vl" ~ "surv | con | drug |vl",
        cohorts == "any_s" ~ "surv",
        cohorts == "any_c" ~ "con",
        cohorts == "any_d" ~ "drug",
        cohorts == "any_v" ~ "vl",
        cohorts == "cond_and_drug" ~ "con & drug",
        cohorts == "svy_and_cond" ~ "surv & con",
        cohorts == "svy_and_drug" ~ "surv & drug",
        cohorts == "vl_and_svy" ~ "vl & surv",
        cohorts == "vl_and_cond" ~ "vl & con",
        cohorts == "vl_and_drug" ~ "vl & drug",
        TRUE ~ "extra"
    )) %>%
    mutate(cohort_name = factor(cohort_name,
                           levels = c(
                               "national",
                               "surv & con & drug & vl",
                               "surv & con & drug | vl",
                               "surv & vl & con | drug",
                               "drug & vl & surv | con",
                               "surv & con | drug",
                               "vl & con | drug",
                               "vl & surv | drug",
                               "vl & surv | con",
                               "surv | con | drug |vl",
                               "surv",
                               "con",
                               "drug",
                               "vl",
                               "con & drug",
                               "surv & con",
                               "surv & drug",
                               "vl & surv",
                               "vl & con",
                               "vl & drug")))
```

#Age

# plot has to combine first two age groups. 
# also remove small N's < 20
```{r}
demo = "age"
df_fig2_age = study_dat %>% 
     filter(variable == !!demo) %>% droplevels() %>%
     mutate(collapse = case_when(
         str_detect(cat, "13|25") ~ 0,
         str_detect(cat, "35") ~ 1,
         str_detect(cat, "45") ~ 2,
         str_detect(cat, "55") ~ 3
         )
            )%>%
    group_by(collapse, cohort_name) %>%
    mutate(
        n = sum(n), 
        total = mean(total)) %>%
    mutate(
        prop = n/total,
        cat = ifelse(cat == "Age 25-34", "Age 13-34", cat)) %>%
    filter(cat != "Age 13-24") %>% droplevels() %>%
     drop_na(cat) %>%
     group_by(cohort_name) %>%
     complete(cat, fill = list(prop = 0, variable = demo)) %>%
      mutate(prop = ifelse(n<20, NA, prop))
# options(repr.plot.width = 16, repr.plot.height = 6, repr.plot.res = 200)
# 
# # cleaned up plot for poster. 
# df_fig2_age %>%
#   ggplot(aes(x = cohort_name, y = prop, fill = cohort_name)) + #, color = color_outline
#     geom_col(position = "dodge", show.legend = FALSE,
#              aes(color = ifelse(cohort_name == "national", "black", "white"))) +
#     scale_color_identity() +
#    facet_wrap(~ cat, scales = "free_x", ncol = 5) + #  ylim(0, 1) +
#    theme(axis.text.x = element_blank(),
#          plot.title = element_text(size = 36, face = "bold", margin = margin(t = 0, r = 15, b = 10, l = 0)),
#      axis.text = element_blank(),
#     axis.title.y = element_text(size = 20, margin = margin(t = 0, r = 20, b = 0, l = 0)),
#      strip.text = element_text(size = 30)) +
#  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
#  labs(title = paste("Age"),
#             x = NULL,
#             y = "Proportion of Participants") + 
#   force_panelsizes(cols = unit("9", "cm"), total_height = unit("12", "cm"), TRUE)
```

# Race

# combine other/skip for race...end up not doing this, but it was easier just to filter it out below
```{r}
demo = "race"
df_fig2_race = study_dat %>% 
    filter(variable == !!demo) %>% droplevels() %>%
    mutate(cat = ifelse(str_detect(cat,"Other|Skip"), "Other/Skip", cat)) %>%
    group_by(cat, cohort_name) %>%
    mutate(
        n = sum(n), 
        total = mean(total),
        prop = n/total) %>%
       # cat = ifelse(cat == "Age 25-34", "Age 13-34", cat)) %>%
     distinct() %>%
     drop_na(cat) %>%
     group_by(cohort_name) %>%
     complete(cat, fill = list(prop = 0, variable = demo)) %>%
     mutate(cat = factor(cat, levels = c("Black", "White", "Hispanic/Latino", "Mixed", "Other/Skip"))) %>%
    mutate(prop = ifelse(n<20, NA, prop))
# df_fig2_race %>%
#     filter(cat != "Other/Skip") %>%
#   ggplot(aes(x = cohort_name, y = prop, fill = cohort_name)) + #, color = color_outline
#     geom_col(position = "dodge", show.legend = FALSE,
#              aes(color = ifelse(cohort_name == "national", "black", "white"))) +
#     scale_color_identity() +
#    facet_wrap(~ cat, scales = "free_x", ncol = 5) +
#    theme(axis.text.x = element_blank(),
#          plot.title = element_text(size = 36, face = "bold", margin = margin(t = 0, r = 15, b = 10, l = 0)),
#      axis.text = element_blank(),
#      #axis.title.x = element_text(size = 14),
#      axis.title.y = element_text(size = 20, margin = margin(t = 0, r = 20, b = 0, l = 0)),
#      #legend.text = element_text(size = 14),
#      #legend.title = element_text(size = 14),
#      #legend.position = "bottom",
#      strip.text = element_text(size = 30)) +
#  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
#  labs(title = paste("Race"),
#             x = NULL,
#             y = "Proportion of Participants") + 
#   force_panelsizes(cols = unit("9", "cm"), total_height = unit("12", "cm"), TRUE)
```


# Gender Sex
```{r}
demo = "gender_sex"
df_fig2_gs = study_dat %>% 
    filter(cat != "Skip") %>%
    filter(variable == !!demo) %>% droplevels() %>%
    #mutate(cat = ifelse(str_detect(cat,"Diverse|Skip")|is.na(cat), "Gender Diverse/Skip", cat)) %>%
    group_by(cat, cohort_name) %>%
    mutate(
        n = sum(n), 
        total = mean(total),
        prop = n/total) %>%
     distinct() %>%
      drop_na(cat) %>%
     group_by(cohort_name) %>%
     complete(cat, fill = list(prop = 0, variable = demo)) %>%
     mutate(cat = factor(cat, levels = c("Male", "Female", "Gender Diverse"))) %>%
    mutate(prop = ifelse(n<20, NA, prop))
# df_fig2_gs %>%
#   filter(cat != "Skip") %>%
#   ggplot(aes(x = cohort_name, y = prop, fill = cohort_name)) + #, color = color_outline
#     geom_col(position = "dodge", show.legend = FALSE,
#              aes(color = ifelse(cohort_name == "national", "black", "white"))) +
#     scale_color_identity() +
#    facet_wrap(~ cat, scales = "free_x", ncol = 5) +
#    theme(axis.text.x = element_blank(),
#          plot.title = element_text(size = 36, face = "bold", margin = margin(t = 0, r = 15, b = 10, l = 0)),
#      axis.text = element_blank(),
#      #axis.title.x = element_text(size = 14),
#      axis.title.y = element_text(size = 20, margin = margin(t = 0, r = 20, b = 0, l = 0)),
#      #legend.text = element_text(size = 14),
#      #legend.title = element_text(size = 14),
#      #legend.position = "bottom",
#      strip.text = element_text(size = 28)) +
#  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
#  labs(title = paste("Gender Sex"),
#             x = NULL,
#             y = "Proportion of Participants") + 
#   force_panelsizes(cols = unit("8", "cm"), total_height = unit("12", "cm"), TRUE)
```

# Table 1

to make table 1, we have to adjust some of the level combinations form above
# then we'll combine the data that went into the plots
# and feed it to the gt package to make a nice table we can save as a word doc to put it directly in the poster
```{r}
demo = "gender_sex"
df_fig2_gs = study_dat %>% 
   # filter(cat != "Skip") %>%
    filter(variable == !!demo) %>% droplevels() %>%
    mutate(cat = ifelse(is.na(cat), "Skip", cat)) %>%
    group_by(cat, cohort_name) %>%
    mutate(
        n = sum(n), 
        total = mean(total),
        prop = n/total) %>%
     distinct() %>%
    #  drop_na(cat) %>%
     group_by(cohort_name) %>%
     complete(cat, fill = list(prop = 0, variable = demo)) %>%
     mutate(cat = factor(cat, levels = c("Male", "Female", "Gender Diverse", "Skip"))) %>%
    mutate(prop = ifelse(n<20, NA, prop))

demo = "race"
df_fig2_race = study_dat %>% 
    filter(variable == !!demo) %>% droplevels() %>%
    mutate(cat = ifelse(str_detect(cat,"Other|Skip"), "Other/Skip", cat)) %>%
    group_by(cat, cohort_name) %>%
    mutate(
        n = sum(n), 
        total = mean(total),
        prop = n/total) %>%
       # cat = ifelse(cat == "Age 25-34", "Age 13-34", cat)) %>%
     distinct() %>%
     drop_na(cat) %>%
     group_by(cohort_name) %>%
     complete(cat, fill = list(prop = 0, variable = demo)) %>%
     mutate(cat = factor(cat, levels = c("Black", "White", "Hispanic/Latino", "Mixed", "Other/Skip"))) %>%
    mutate(prop = ifelse(n<20, NA, prop))
```


``` {r}
t1 = bind_rows(
    df_fig2_age, 
    df_fig2_race,
    df_fig2_gs
    ) %>%
    select(-collapse) %>%
    mutate(cell_value = ifelse(!is.na(n)& n>=20, paste0(n, " (", round(prop,2)*100, ")"), "<20")) %>%
    select(-total, -n, -prop) %>%
    filter(cohorts != "nat") %>% arrange(variable, cohort_name, cat) 

t2 = denominators %>%
    filter(cohorts != "nat") %>%
            mutate(
                  cohorts = case_when(
                      cohorts == "svy" ~ "any_s",
                      cohorts == "cond" ~ "any_c",
                      cohorts == "drug" ~ "any_d",
                      cohorts == "vl" ~ "any_v",
                      TRUE ~ cohorts
                  ),
                  cell_value = as.character(total),
                  variable = "cohort total",
                  cat = "total") %>% select(-total) %>%
    filter(cohorts %in% c("nat", "scdv", "svy_and_cond_and_drug_or_vl", "svy_and_vl_and_cond_or_drug", "drug_and_vl_and_svy_or_cond",
                          "svy_and_cond_or_drug", "vl_and_cond_or_drug", "vl_and_svy_or_drug", "vl_and_svy_or_cond",
                          "svy_or_cond_or_drug_or_vl", "any_s", "any_c", "any_d", "any_v",
                          "cond_and_drug", "svy_and_cond", "svy_and_drug", "vl_and_svy", "vl_and_cond", "vl_and_drug")) %>%
    mutate(cohort_name = case_when(
        cohorts == "nat" ~ "national",
        cohorts == "scdv" ~ "surv & con & drug & vl",
        cohorts == "svy_and_cond_and_drug_or_vl" ~ "surv & con & drug | vl",
        cohorts == "svy_and_vl_and_cond_or_drug" ~ "surv & vl & con | drug",
        cohorts == "drug_and_vl_and_svy_or_cond" ~ "drug & vl & surv | con",
        cohorts == "svy_and_cond_or_drug" ~ "surv & con | drug",
        cohorts == "vl_and_cond_or_drug" ~ "vl & con | drug",
        cohorts == "vl_and_svy_or_drug" ~ "vl & surv | drug",
        cohorts == "vl_and_svy_or_cond" ~ "vl & surv | con",
        cohorts == "svy_or_cond_or_drug_or_vl" ~ "surv | con | drug |vl",
        cohorts == "any_s" ~ "surv",
        cohorts == "any_c" ~ "con",
        cohorts == "any_d" ~ "drug",
        cohorts == "any_v" ~ "vl",
        cohorts == "cond_and_drug" ~ "con & drug",
        cohorts == "svy_and_cond" ~ "surv & con",
        cohorts == "svy_and_drug" ~ "surv & drug",
        cohorts == "vl_and_svy" ~ "vl & surv",
        cohorts == "vl_and_cond" ~ "vl & con",
        cohorts == "vl_and_drug" ~ "vl & drug",
        TRUE ~ "extra"
    )) %>%
    mutate(cohort_name = factor(cohort_name,
                           levels = c(
                               "national",
                               "surv & con & drug & vl",
                               "surv & con & drug | vl",
                               "surv & vl & con | drug",
                               "drug & vl & surv | con",
                               "surv & con | drug",
                               "vl & con | drug",
                               "vl & surv | drug",
                               "vl & surv | con",
                               "surv | con | drug |vl",
                               "surv",
                               "con",
                               "drug",
                               "vl",
                               "con & drug",
                               "surv & con",
                               "surv & drug",
                               "vl & surv",
                               "vl & con",
                               "vl & drug"
                           ))) %>%
     bind_rows(t1) %>%
     select(-cohorts) %>%
     pivot_wider(names_from = "cohort_name", values_from = cell_value) %>%
     select(variable, cat, 'surv & con & drug & vl', 'surv & con & drug | vl', 'surv & vl & con | drug', 'drug & vl & surv | con',
            'surv & con | drug', 'vl & con | drug', 'vl & surv | drug', 'vl & surv | con', 'surv | con | drug |vl', 'surv',
            'con', 'drug', 'vl', 'con & drug', 'surv & con', 'surv & drug', 'vl & surv', 'vl & con', 'vl & drug' ) %>%
    mutate(order = c(0, 1, 2, 3, 4, 5, 7, 8, 6, 8.5, 11, 9, 10, 12)) %>% arrange(order) %>% select(-order)
    
#### add back 15 after 14, only if the 45-54, 55+ goups are added back

# preview
t2

# fix for trying to do gt in jupyter notebooks
gts <- function(gt_table){
   gt:::as.tags.gt_tbl(gt_table)
}

t2 %>% 
    gt(groupname_col = "variable") %>%
    gts()
```


